\input cheatmac
\usepackage[czech]{babel}
\begin{document}
\begin{multicols}{3}

\title{Probabilistic algorithms}

\fsection{Basic algorithms}

{\sc Mincut:} Contract $n-2$ edges (keep parallel, delete loops),
return the rest as a cut.

\thm{$P[$resulting cut minimal$] ≥ {2 \over n(n-1)}$.}

\prf{Suppose min. cut has $k$ edges. $||G|| ≥ {nk \over 2}$.\hfil\break
$P[E_1 ≡$ first choice misses min. cut$] ≥ 1 - {k \over nk/2} = 1 - {2 \over n}$.\hfil\break
$P[E_i |$ missed all until now $] ≥ 1 - { 2 \over n-1}$.\hfil\break
$P[$we missed all $] ≥ ∏_{i=1}^{n-2}( 1- {2 \over n+1-i}) = {2 \over n(n-1)}$.
Repeat $n^2 \ln n$ times for $1 \over n^2$ probability of missing.
}

{\sc FastSelect:} $R ≡ $ sample $⌈n^{3/4}⌉$ elements, sort. \hfil\break
Pick $d ≡ R[⌋n^{3/4}/2 - \sqrt{n}⌊]$, $u ≡ R[⌈ n^{3/4}/2 + \sqrt{n}⌉]$.
\begin{itemize}
\item $|R[1..d]| > n/2$,
\item $|R[u..]| > n/2$,
\item $|R[d..u]| ≥ 4n^{3/4}$.
\end{itemize}
Sort $R[d..u]$, return median.

\obs{$P[(a)] ≤ 1/4 n^{-1/4}$.}

\prf{ $X_i ≡ $ chosen element is $≤$ median. $P[X_i = 1] = 1/2 + {1 \over 2n}$.
$Var[∑_{i=1}^{n^{3/4}}] = n^{3/4} (1/2 + {1 \over 2n}) (1/2 - {1 \over 2n}) = 1/4 n^{3/4} - {n^{3/4} \over 4n^2} < 1/4 n^{3/4}$.

Chebyshev: $P[∑ X_i < 1/2 n^{3/4} - \sqrt{n}] ≤ P[|∑ X_i - E[∑ X_i]| > \sqrt(n} ≤ {Var[∑ X_i] \over n}] = 1/4 n^{-1/4}$.
% }

\obs{ $P[(c)] ≤ 1/2 n^{-1/4}$.}

\prf{ Either at least $2n^{3/4}$ elements are above (or below) median. Apply binomial distro again + Chebyshev.}

{\sc Hypercube Routing:} 

\thm{Markov} $X$ nezáporná, $a > 0$:
$P[X > \alpha E[X]] < {1 \over \alpha}.$

\thm{Čebyšev}
$P[|X-E[X]| \ge \lambda\sigma] \le {1 \over \lambda^2}.$

\prf{}
$\sigma^2 = \var[X] = E[(X - E[X])^2] \ge \lambda^2 \sigma^2 P[|X-E[X]| \ge \lambda\sigma. $

\thm{Černov} Nechť $S_n$ je součtová náhodná proměnná pro $n$ mincí s hodnotami ${+1,-1}$. Pak:
$P[S_n > a] < e^{-a^2 / 2n}.$

\prf{} Vytvoříme náhodnou proměnnou $Y = e^{uX}$.

$E[Y] = \prod_{i=1}^n E[e^{uX_i}] = {e^u + e^{-u} \over 2}^n \le e^{nu^2/2}$. Použitím Markova pak
$P[X \ge t] = P[Y \ge e^{ut}] \le E[Y] / e^{ut} \le e^{nu^2/2 - ut}$. Zvol $u =
t/n$.


\fsection{Random walks}
In the general setting, we're throwing a $d(v)$ sided dice.

\dfn{} $G$ leads to a transition matrix $P$; $p_{uv} = 1/d(u)$ if $uv \in E$, zero otherwise.

\dfn{} $h_{uv}$ is hitting time of a pair $(u,v)$. Cover time $C(G) \equiv$ hitting all edges.

\thm{} Stationary distribution of a r.w. on $G$ is $\pi_{u} = d(u)/2|E|$. 

\thm{} $uv \in E \then h_{uv} + h_{vu} \le 2|E|$. Equality if $uv$ bridge.

\thm{} $C(G) \le  2m(n-1) < 2n^3$. 

\prf{} Pick a spanning tree, count hitting times along it.

\dfn{} $R_uv \equiv$ resistance along $uv$ if we replace all edges with $1\Omega$ resistors.

\thm{} $h_uv + h_vu = 2|E|R_{uv}$.

\prf{} Put $d(z)$ current into each vertex, collect $2m$ at $v$.
$\Phi_{zv} \equiv$ voltage between $z$ and $v$. Voltage is voltage, so:
$\forall z \ne v:  \sum_{w|zw \in E}(\Phi_{zv} - \Phi_{wv}) = d(z)$. From this:
$\Phi_{zv} = 1 + \sum {\Phi_{zw} \over d(z)}$, which is the same equation as
for $h_uv$.

Now reverse the flow of the current, you get $h_vu$. Sum the two currents and
use $U=IR$ and done.

{\bf 2-SAT:} choose a random assignment, then pick an unsatisfied clause and
change a random variable within. Probability of moving forward $\ge 1/2 \then
n^2$ expected time until we hit the goal.

{\bf 3-SAT:} in NP, so probabilistic polynomial algorithm unlikely. $2^n$: try all.
Better: choose a random assignment, try a short random walk, reset.

\obs{ $P[$ hit the goal in $3k$ steps $| k$ steps away $ ] \ge
{n \choose 2k}$ ...}

\dfn{ {\tt RL} $\equiv$ class of languages recognizable in space $\O(\log n)$ and polyn. time,
one-sided error.}

\thm{ $USTCON \in {\tt RL}$.}

\prf{ $4n^3$ steps + Markov + $c(G)$ bound.}

\thm[Reingold]{ $USTCON \in {\tt L}$. }

\end{multicols}
\end{document}